Discovering causal factors with back-and-forth regression

### 2019-05-21 paper/paper.tex
wip: intro method

\begin{abstract}
Identifying causes from observations is at the core of science. This endeavor
is particularly challenging when i) potential factors are difficult to
manipulate individually and ii) observations are complex and multi-dimensional.
To address this issue, we introduce ``back-and-forth regression'' (BFR), a
method that optimally identifies the most plausible causal factors from a set
of covarying candidates. BFR consists of i) predicting the potential causes $X$
from the observable $Y$, and subsequently ii) identifying the only factors that
contributed to those predictions. After detailing the proof of convergence, we
show that BFR outperforms least-squares and related regression techniques, as
well as cross-decomposition methods (e.g. canonical correlation analysis, and
partial least squares) on two tasks: causal identification and out-of-distribution
prediction. We apply BFR to several empirical datasets to show that it
retrieves plausible results. In particular, we apply BFR to neuroimaging
recordings of 102 subjects presented with word sequences, and show, as
expected, that the early brain responses appear to be specifically caused by
low-level visual factors whereas late brain responses appear to be caused by
lexico-semantic factors, despite the fact that these factors covary. We finish
by extending BFR to nonlinear cause-effect relationships.
\end{abstract}

@@

Causal discovery, telling causes from their effects, is an inverse problem.
%
Given experimental data (effects) and possible explanations (causes), we are tasked to find the causes that account for the observations, knowing that other non causal factors (contingent, random) are at work.

More precisely, we measure, via an apparatus F, a multidimensional signal Y, that is a function of possible causes X, and other, unknown, factors that can be understood as noise N: $Y=f(X,N)$.
%
Of all possible causes (features of X), only a few account for Y, let them be represented as EX, with E a binary diagonal matrix: a diagonal of zeroes and ones, which correspond to active causal features of X.
%
Let us assume that noise N is centred (adding, if need be, an intercept to the active causal features, to capture the bias in N), and additive (not a very strong assumption in the absence of further constraints on N).
%
Then, we can write our model of causal discovery as $Y=f(m(EX)+N)$, with f and m unknown functions.
%
Up to some rearrangement of the features of X, we can also suppose m linear.

We end up with the following model $Y=f(MEX+N)$, where N is centred noise E a binary diagonal matrix, M a full rank matrix and f some unknown transformation.
%
Given experimental measures of X and Y (as matrices), and our goal is to recover E, the active causal features of the process under study.

*** description of the neuroscience problem

Such problems have been attacked along two different lines.
%
In the direct approach, we use the above equation as a statistical model, try to express each feature of Y as a function of X, test whether the coefficients associated with specific features of X are different from zero, and recover E as the features of X active over some dimension of Y.
%
In the inverse approach, we try to express each feature of X as a function of Y, estimate its changes as Y varies, and recover E as the set of "non random" features of X.
%
In the presence of noise and a large number of features of X, Y and N, both approaches are limited because they fail to take advantage of the correlations between features of the variable they are modelling (or guard against their multicollinearity).
%
In the direct approach, features of Y are modelled separately.
%
In the inverse approaches, correlations between features of X are not accounted for.

In this paper, we introduce back and forth regression, which combines direct and inverse approaches.
%
We first predict X from Y, as in the inverse approach, and retrieve a robust least square estimate $\hat X$. Then, we calculate a robust least square estimator of $\hat X$ from X.
%
If the active features of X are represented as EX, with E a binary diagonal, this provides a good approximation of E, robust to noise and environment change.
%
The main purpose of back and forth regression is feature extraction, but it can be turned into a regression method by filtering X on the selected features.

In the next sections, we describe back and forth regression, prove that it asymptotically extracts E, both in the linear and non linear case, and that it can be understood as a special case of canonical component analysis (cca).
%
By experimenting with simulated data, we demonstrate its robustness and use cases, and show that it compares favourably with state of the art techniques for regression and feature extraction.
%
Finally, we illustrate its use on an example from neuroscience.

@@

\subsection{The method}

We are given two set of measurements over a sample of N individuals.
%
Y are ny multidimensional effects, and X are nx multidimensional possible causes.
%
We assume that only some features of X influence Y, so that $Y=f(EX,N)$, E being a binary diagonal, N some unknown noise, and f an unknown function.
%
We try to recover E, the active causal features.

@@

Under our hypotheses (centred noise, E a binary diagonal) and even under heavy noise, $\hat H$ is close to diagonal.
%
Yet, because of regularisation which adds positive bias to small eigenvalues, and noise which attenuates large eigenvalues, $\hat H$ is in fact a scaled and noisy approximation to E.
%
To recover a binary diagonal, as our estimate of E, we extract the diagonal of $\hat H$ and select its large elements.
%
The diagonal matrix $\hat E$, having large elements of H as its unit diagonal coefficients (the rest being zero) is our estimate of E, the active features of X.

To select large elements of $\hat H$, the traditional approach in regression analysis consists in testing whether its coefficients differ from zero. This is usually done via variants of Student t-test.
%
This will not work here, as back and forth regression introduces, through two regularisations, a significant amount of positive bias on the diagonal, which depends in a complex manner on noise, dimensionality and conditioning of X and Y.
%
Instead, we treat feature extraction as a one dimensional binary clustering problem: classify n positive real values as "large" and "small" (under the hypothesis that they are not all small or all large).

This can be done by maximising the ratio of inter-group variance over total variance (ie minimising intra-group variance).
%
In a one dimensional two clusters setting, this amounts to sorting the data, and separating the p smallest from the n-p largest values so that inter group inertia is maximal.
%
If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we are maximising the Sonquist and Morgan criterion $$K = {p (n-p) \over n} {(s-r)^2 \over V}$$ and selecting the p features corresponding to the largest values.
%
If diagonal elements of $\hat H$ are normally distributed, K follows a chi-square distribution with one degree of freedom.
%
At 95\% confidence level, a value of K superior to 3.84 makes our selection significant (for 99\% confidence, the corresponding value of K is 6.63) \citep{Kass_75}

This extract from $\hat H$ a binary diagonal $\hat E$.
%
To use back and forth regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.

Summarizing, back and forth regression is performed as follows:
\begin{enumerate}
\item Randomly split the data sample.
\item From the first half sample, using crossalidated ridge regression, find $\hat G$, the regularised least square regression of X from Y: $\hat G=(Y'Y)^{-1} Y'X$.
\item From the second half sample, calculate $\hat X = \hat G Y$, and find $\hat H$, the regularized least square regression of $\hat X$ from X: $\hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X$.
\item Repeat steps 1 to 3 and average the results as $\hat H$.
\item Using the Sonquist and Morgan criterion, select the large diagonal elements of $\hat H$, set them to one and the rest of $\hat  H$ to zero.
\item Use the resulting matrix $\hat E$ as an estimator of E, the active features in the causal process
\item Alternatively, use $\hat E$ to predict Y from X, by performing a regularised least square regression of Y from $\hat E X$.
%

@@

The two steps of back and forth regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).

@@

Thus, in the linear case, back and forth regression retrieves $\hat H = \frac{1}{1+nsr}(FE)^{\dagger}(FE)$.

@@

\section{Related works}
\subsection{back and forth regression as a special case of canonical component analysis}
Formula (1) shows that back and forth regression can be seen as a special case of canonical component analysis (CCA).
%
In CCA, we are given two matrixes X and Y (or size N,p and N,q) that describe different features measured on the same data sample, and characterise their proximity (that is, between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.

This is done by finding vectors $a$ and $b$ such that $Xa$ and $Yb$ display maximal correlation, or by maximising $a'X'Yb$ over unit-normed $a$ and $b$.
%
To find a, one calculates the eigenvectors of (1).
%
The largest one is the direction with maximal correlation, the second one the maximal correlation direction orthogonal to the first, and so on.
%
For each canonical dimension, the corresponding eigenvalue is the square of the cosine between the corresponding direction along X and the subspace spanned by Y.

If (1) is diagonal, for instance in the model presented above, the features of X are eigenvectors.
%
If all eigenvalues are either 0 or 1, since they correspond to the squared cosine of the angle between the corresponding feature and the subspace spanned by Y, all the features are either part of the Y subspace (eigenvalues of one) or orthogonal to it (for zero eigenvalues).
%
Thus, back and forth regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
%
The closer we are to this hypothesis, the closer the results of CCA and back and forth regression will be.

There are differences between CCA and back and forth regression.
%
First, since our objective is not to characterise the proximity between X and Y, but just to extract causal features, we can dispense with the last step of CCA (diagonalisation of (1)).
%
Second, since we operate in a noisy environment, we use regularisation, introducing bias in the correlation calculations.
%
Finally, whereas CCA wants to leverage all correlations between X and Y, we filter some of them, through splitting and bagging.
%
In other words, whereas we are working from the same formulae as CCA, we use them in different ways, for different purposes

@@

\section{Experiments - simulated data}
Experiments on simulated data serve three purposes: compare back and forth regression with other known techniques (CCA, OLS, PLS...) both as a regression and a feature extraction method, measure its performance for various values of its parameters and provide evidence for our claims about stability, invariance to environment, and behavior as a causal detector.

@@

We evaluate the performance of back and forth regression in two different roles : as a feature extractor recovering E, and as a regression method, predicting Y from X over a test sample.
%
For feature extraction, the quality of fit is the number of disagreements between our estimator and the diagonal of E (ie, their Hamming distance, which should be minimised).
%
For regression, we use a test set of N additional measurements, and calculate the R-squared ratio of variances (which should be maximised).
%
Performance indicators are calculated by averaging the results of several experiments.

\subsection{Feature extraction - linear case}
All other parameters being equal, the quality of feature extraction, measured as the discrepancy between features extracted by back and forth regression ($\hat E$) and the diagonal of E, improves with signal to noise ratio.
%
The decrease is usually very steep until a Hamming distance under 2 is achieved.
%
Figure 1 presents a number of such curves for varying values of dx=dy=dz, nc=30 and N=1000.

@@

For feature extraction, back and forth regression either works very well (Hamming distance less than 2) or very badly, with a sharp change over small variations of the parameters.
%
The following graphs show the boundary between values of dx and dz that allow for features extractions (dx lower than boundary, or dz larger), for different values of dy (level curves on the graph), and different values of signal to noise ratio and number of active variables.

@@

For values of ...
%
in ..., an empirical formula can be provided.
%
back and forth regression allows for good feature extraction if :

@@

\subsection{Regression }
For regression, we want to understand how feature extraction helps predict Y from X.
%
To this effect, we compare four approaches to regression:
\begin{itemize}
\item standard ridge regression of Y from X, where no features are extracted this is our baseline
\item weighted ridge regression, of Y from $\hat H X$, where features of X are weighted according to the diagonal of $\hat H$
\item filtered ridge regression, of Y from $\hat E X$, where active features are first extracted (as above), and Y is predicted from the extracted features only
\item perfect filtered (oracle) ridge regression of Y from EX, where E is supposed known
\end{itemize}
Over a large set of parameters, we observe that filtered regression is on average better than weighted, which is on average better than standard regression.
%
This demonstrates that back and forth regression can be used as a regression method, over the areas in parameter space where feature extraction performs well.

@@

The rationale behind such tests is that whereas back and forth regression, being a feature extraction method, might not be as efficient as dedicated prediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.

@@

@@

>>>>

\begin{abstract}
Identifying causes from observations is at the core of science. This endeavor
is particularly challenging when i) potential factors are difficult to
manipulate individually and ii) observations are complex and multi-dimensional.
To address this issue, we introduce ``back-and-forth regression'' (BFR), a
method that optimally identifies the most plausible causal factors from a set
of covarying candidates. BFR consists in i) predicting the potential causes $X$
from the observable $Y$, and subsequently ii) identifying the only factors that
contributed to those predictions. After detailing the proof of convergence, we
show that BFR outperforms least-squares and related regression techniques, as
well as cross-decomposition methods (e.g. canonical correlation analysis, and
partial least squares) on two tasks: causal identification and out-of-distribution
prediction. We apply BFR to neuroimaging recordings of 102 subjects reading word
sequences. The results show, as expected, that the early brain responses to
words appear to be specifically caused by low-level visual factors whereas late
brain responses appear to be caused by lexical factors, despite the fact that
these factors covary in the study. We finish by extending BFR to nonlinear cause-effect relationships.
\end{abstract}

@@

Natural sciences are tasked to find, from a set of potentially-covarying factors, the minimal subset that suffices to predict a set of potentially multidimensional observations.
%
This problem can be formalized as follow: we measure, via an apparatus F, a multidimensional signal Y, that is a function of possible causes X, and other unknown factors that can be understood as noise N: $Y=f(X,N)$.
%
Of all possible causes (features of X), only a subset may account for Y, let them be represented as EX, with E a binary diagonal matrix: a diagonal of zeroes and ones, which correspond to active causal features of X.
%
Under the assumption of centered and additive noise N, causal discovery thus consists in finding E in $Y=f(EX+N)$, with f and N being unknown.

%
Causal discovery is impeded by two challenges.
%
First, causal and non-causal factors may naturally co-vary. [For example, XXX.] (e.g. X'X is not an orthogonal matrix). When the number of co-varying factors is low, this issue is generally solved with factorial designs [cite the design of experiments, fisher 1935]. [XXX medical example]. As factorial design become hard to implement with numerous factors, the orthogonalization of co-varying factors is performed analytically, by estimating how each factor contributes to the variance of the observations (e.g analysis of variance, ANOVA). In this paper, we refer to this approach as 'forward modeling' as it models how factors X predict the observations Y. As illustrated in Fig X., forward model can be sub-optimal when observations are multidimensional and noisy, because it typically does not make use of the structure of observations (e.g. the model does not consider Y'Y).
%
Second, causal discovery can be challenged by the multiplicity and complexity of the observations. [For example XXX]. This issue has led many to reverse the problem and to focus on optimally predicting factors from complex and poorly controlled observations. In neuroimaging, for example, it is increasingly popular to use "decoding" methods: instead of testing whether the brain responds to particular factors, mathematical models are trained to maximally predict such factors from complex brain signals. We refer to this approach as 'backward modeling' as it models how the observations Y can be used to predict the factors X. However, backward can predict a non-causal factor as long as it correlates with a causal one.

%
In sum, forward modeling approaches are well suited in disentangling the independent contribution of covarying factors, but can be suboptimal at combining multidimensional observations. By contrast, backward modeling approaches are well suited at combining multiple observations, but do not disentangle covarying factors [Figure X].
%
In this paper, we introduce the 'back-and-forth regression', which combines the benefit of forward and backward approaches. Back-and-forth regression consists in first predicting X from Y (as in the backward modeling) subsequently retrieving a robust least square estimate $\hat X$ and finally computing a least square estimator of $\hat X$ from X (as in forward modeling).
%
If the active features of X are represented as EX, with E a binary diagonal, this provides a good approximation of E, robust to noise and environment change.
%
The main purpose of back-and-forth regression is feature extraction, but it can be turned into a regression method by filtering X on the selected features.

In the next sections, we (i) describe back-and-forth regression, (ii) prove that it asymptotically extracts E, (iii) both in the linear and non linear case, and that it can be understood as a special case of canonical component analysis (cca).
%
By experimenting with simulated data, we demonstrate the robustness and typical use cases of back-and-forth regression, and show that it compares favourably with state of the art techniques for regression and feature extraction, both in terms of causal discovery and out-of-
%
Finally, we illustrate its use on an example from neuroimaging.

@@

\subsection{Algorithm}

We are given two set of measurements over a sample of n individuals.
%
% Francois can you use the Y in Real^{n, ny} format
%
Y are ny multidimensional effects, and X are nx multidimensional possible causes.
%
We assume that only some features of X influence Y, so that $Y=f(EX,N)$, E being a binary diagonal, N some unknown noise, and f an unknown function.
%
We try to recover E, the active causal features.

@@

Under our hypotheses (centered noise, E a binary diagonal) and even under heavy noise, $\hat H$ is close to diagonal.
%
Yet, because of regularisation which adds positive bias to small eigenvalues, and noise which attenuates large eigenvalues, $\hat H$ is in fact a scaled and noisy approximation to E.
%
To recover a binary diagonal, as our estimate of E, we extract the diagonal of $\hat H$ and select its large elements.
%
The diagonal matrix $\hat E$, having large elements of H as its unit diagonal coefficients (the rest being zero) is our estimate of E, the active features of X.

To select large elements of $\hat H$, the traditional approach in regression analysis consists in testing whether its coefficients differ from zero. This is usually done via variants of Student t-test.
%
This will not work here, as back-and-forth regression introduces, through two regularisations, a significant amount of positive bias on the diagonal, which depends in a complex manner on noise, dimensionality and conditioning of X and Y.
%
Instead, we treat feature extraction as a one dimensional binary clustering problem: classify n positive real values as "large" and "small" (under the hypothesis that they are not all small or all large).

This can be done by maximizing the ratio of inter-group variance over total variance (ie minimizing intra-group variance).
%
In a one dimensional two clusters setting, this amounts to sorting the data, and separating the p smallest from the n-p largest values so that inter group inertia is maximal.
%
If s and r are the average values of the two clusters, p and n-p their size and V the total variance of the sample, we are maximising the Sonquist and Morgan criterion $$K = {p (n-p) \over n} {(s-r)^2 \over V}$$ and selecting the p features corresponding to the largest values.
%
If diagonal elements of $\hat H$ are normally distributed, K follows a chi-square distribution with one degree of freedom.
%
At 95\% confidence level, a value of K superior to 3.84 makes our selection significant (for 99\% confidence, the corresponding value of K is 6.63) \citep{Kass_75}

This extract from $\hat H$ a binary diagonal $\hat E$.
%
To use back-and-forth regression as a regression technique (to predict Y from X), we multiply X by $\hat E$ and perform a cross validated ridge regression of each feature of Y from $\hat E X$.

Summarizing, back-and-forth regression is performed as follows:
\begin{enumerate}
\item Randomly split the data sample.
\item From the first half sample, using crossalidated ridge regression, find $\hat G$, the regularised least square regression of X from Y: $\hat G=(Y'Y)^{-1} Y'X$.
\item From the second half sample, calculate $\hat X = \hat G Y$, and find $\hat H$, the regularized least square regression of $\hat X$ from X: $\hat H=(X'X)^{-1} X'Y(Y'Y)^{-1} Y'X$.
\item Repeat steps 1 to 3 and average the results as $\hat H$.
\item Using the Sonquist and Morgan criterion, select the large diagonal elements of $\hat H$, set them to one and the rest of $\hat  H$ to zero.
\item Use the resulting matrix $\hat E$ as an estimator of E, the active features in the causal process
\item Alternatively, use $\hat E$ to predict Y from X, by performing a regularised least square regression of Y from $\hat E X$.
%

@@

The two steps of back-and-forth regression amounts to finding $\hat G=\arg \min_G \left \| X-GY \right \|^2$ and $\hat H =\arg \min_H \left \| \hat GY - HX \right \|^2$ (all matrix norms henceforth are Frobenius norms).

@@

Thus, in the linear case, back-and-forth regression retrieves $\hat H = \frac{1}{1+nsr}(FE)^{\dagger}(FE)$.

@@

\section{Related works}
\subsection{back-and-forth regression as a special case of canonical component analysis?}
Formula (1) shows that back-and-forth regression may be related to canonical component analysis (CCA).
%
In CCA, we are given two matrices X and Y (or size N,p and N,q) that describe different features measured on the same data sample, and characterize their proximity (that is, between the subspaces spanned by their columns) as the correlation between linear combinations of X and Y.

This is done by finding vectors $a$ and $b$ such that $Xa$ and $Yb$ display maximal correlation, or by maximizing $a'X'Yb$ over unit-normed $a$ and $b$.
%
To find a, one calculates the eigenvectors of (1).
%
The largest one is the direction with maximal correlation, the second one the maximal correlation direction orthogonal to the first, and so on.
%
For each canonical dimension, the corresponding eigenvalue is the square of the cosine between the corresponding direction along X and the subspace spanned by Y.

If (1) is diagonal, for instance in the model presented above, the features of X are eigenvectors.
%
If all eigenvalues are either 0 or 1, since they correspond to the squared cosine of the angle between the corresponding feature and the subspace spanned by Y, all the features are either part of the Y subspace (eigenvalues of one) or orthogonal to it (for zero eigenvalues).
%
Thus, back-and-forth regression amounts to CCA under the constraint that the subspaces spanned by X and Y are perpendicular, X is spanned by vectors of Y and vectors orthogonal to Y, and that the parallel and orthogonal components of X are spanned by disjoint sets of features of X.
%
The closer we are to this hypothesis, the closer the results of CCA and back-and-forth regression will be.

There are differences between CCA and back-and-forth regression.
%
First, since our objective is not to characterise the proximity between X and Y, but just to extract causal features, we can dispense with the last step of CCA (diagonalisation of (1)).
%
Second, since we operate in a noisy environment, we use regularisation, introducing bias in the correlation calculations.
%
Finally, whereas CCA wants to leverage all correlations between X and Y, we filter some of them, through splitting and bagging.
%
In other words, whereas we are working from the same formulae as CCA, we use them in different ways, for different purposes

@@

\section{Experiments - simulated data}
Experiments on simulated data serve three purposes: compare back-and-forth regression with other known techniques (CCA, OLS, PLS...) both as a regression and a feature extraction method, measure its performance for various values of its parameters and provide evidence for our claims about stability, invariance to environment, and behavior as a causal detector.

@@

We evaluate the performance of back-and-forth regression in two different roles : as a feature extractor recovering E, and as a regression method, predicting Y from X over a test sample.
%
For feature extraction, the quality of fit is the number of disagreements between our estimator and the diagonal of E (ie, their Hamming distance, which should be minimised).
%
For regression, we use a test set of N additional measurements, and calculate the R-squared ratio of variances (which should be maximised).
%
Performance indicators are calculated by averaging the results of several experiments.

\subsection{Feature extraction - linear case}
All other parameters being equal, the quality of feature extraction, measured as the discrepancy between features extracted by back-and-forth regression ($\hat E$) and the diagonal of E, improves with signal to noise ratio.
%
The decrease is usually very steep until a Hamming distance under 2 is achieved.
%
Figure 1 presents a number of such curves for varying values of dx=dy=dz, nc=30 and N=1000.

@@

For feature extraction, back-and-forth regression either works very well (Hamming distance less than 2) or very badly, with a sharp change over small variations of the parameters.
%
The following graphs show the boundary between values of dx and dz that allow for features extractions (dx lower than boundary, or dz larger), for different values of dy (level curves on the graph), and different values of signal to noise ratio and number of active variables.

@@

For values of ...
%
in ..., an empirical formula can be provided.
%
back-and-forth regression allows for good feature extraction if :

@@

\subsection{Regression }
For regression, we want to understand how feature extraction helps predict Y from X.
%
To this effect, we compare four approaches to regression:
\begin{itemize}
\item standard ridge regression of Y from X, where no features are extracted this is our baseline
\item weighted ridge regression, of Y from $\hat H X$, where features of X are weighted according to the diagonal of $\hat H$
\item filtered ridge regression, of Y from $\hat E X$, where active features are first extracted (as above), and Y is predicted from the extracted features only
\item perfect filtered (oracle) ridge regression of Y from EX, where E is supposed known
\end{itemize}
Over a large set of parameters, we observe that filtered regression is on average better than weighted, which is on average better than standard regression.
%
This demonstrates that back-and-forth regression can be used as a regression method, over the areas in parameter space where feature extraction performs well.

@@

The rationale behind such tests is that whereas back-and-forth regression, being a feature extraction method, might not be as efficient as dedicated prediction tools, its focus on causal elements (and the removal of non causal features before carrying out regression) might cause it to be more resistant to changes in environments.

@@

\input{meg_exps.tex}

@@

\section{Acknowlegements}
Data were provided (in part) by the Donders Institute for Brain, Cognition and Behaviour.
